### 2024-07-02 causal-inference.qmd
20240702 - confounding and mediation

```{r}
dagitty::impliedConditionalIndependencies(confounding)

dagitty::adjustmentSets(
  confounding, 
  exposure = "x",
  outcome = "y",
  effect = "total")
```

@@

```{r}
dagitty::impliedConditionalIndependencies(full_mediation)

dagitty::adjustmentSets(
  full_mediation, 
  exposure = "x",
  outcome = "y",
  effect = "total")

dagitty::adjustmentSets(
  full_mediation, 
  exposure = "x",
  outcome = "y",
  effect = "direct")
```

>>>>

```{r}
dagitty::impliedConditionalIndependencies(confounding)
```

The output indicates that player endurance (`X`) and field goals made (`Y`) are conditionally independent when accounting for stadium altitude (`Z`).

```{r}
dagitty::adjustmentSets(
  confounding, 
  exposure = "x",
  outcome = "y",
  effect = "total")
```

The output indicates that, to obtain an unbiased estimate of the causal association between two variables, it is necessary to control for any confounding [@Lederer2019].
That is, to obtain an unbiased estimate of the causal association between player endurance (`X`) and field goals made (`Y`), it is necessary to control for stadium altitude (`Z`).

@@

```{r}
dagitty::impliedConditionalIndependencies(full_mediation)
```

In full mediation, `X` and `Y` are conditionally independent when accounting for the mediator (`Z`).
In this case, coaching quality (`X`) and fantasy points (`Y`) are conditionally independent when accounting for player preparation (`M`).
In other words, in this example, player preparation is the mechanism that fully (i.e., 100%) accounts for the effect of coaching quality on players' fantasy points.

```{r}
dagitty::adjustmentSets(
  full_mediation, 
  exposure = "x",
  outcome = "y",
  effect = "direct")
```

The output indicates that, to obtain an unbiased estimate of the *direct* causal association between coaching quality (`X`) and fantasy points (`Y`) (i.e., the effect that is *not* mediated through intermediate processes), it is necessary to control for player preparation (`M`).

```{r}
dagitty::adjustmentSets(
  full_mediation, 
  exposure = "x",
  outcome = "y",
  effect = "total")
```

However, to obtain an unbiased estimate of the *total* causal association between coaching quality (`X`) and fantasy points (`Y`) (i.e., including the portion of the effect that is mediated through intermediate processes), it is important *not* to control for player preparation (`M`).
When the goal is to understand the (total) causal effect of coaching quality (`X`) and fantasy points (`Y`), controlling for the mediator (player preparation; `M`) would be inappropriate because doing so would remove the causal effect, thus artificially reducing the estimate of the association between coaching quality (`X`) and fantasy points (`Y`) [@Lederer2019].
